---
title: "Code to reproduce the analyses of the AWST paper: Analysis of the LGG data"
author: "[S.M.Pagnotta](https://www.researchgate.net/profile/Stefano_Pagnotta) - Dept. of Science and Technology, Universita' degli Studi del Sannio (Benevento, Italy)</br> [D.Risso](https://www.researchgate.net/profile/Davide_Risso) - Dept. of Statistical Sciences, Universita' degli Studi di Padova (Padova, Italy)"
date:  'compiled on `r format(Sys.time(), "%B %d, %Y")`'
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    toc_float: yes
    df_print: paged
fontsize: 11pt
editor_options: 
  chunk_output_type: console
---

```{r knitr setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=7, fig.align = "center") 
```

Here we show the code to reproduce the analyses of: Risso and Pagnotta (2019). Within-sample standardization and asymmetric winsorization lead to accurate classification of RNA-seq expression profiles. In preparation.

This file belongs to the repository: https://github.com/drisso/awst_analysis.

The code is released with license GPL v3.0.

# Install and load `awst`

```{r install_awst, eval=FALSE, echo=TRUE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("drisso/awst")
```

# Data import and cleaning

For this analysis, we used the collection of 516 samples from the TCGA legacy repository, specifically the RSEM estimates of 20{,}531 gene expressions.

The level-3 LGG data are available in the [Genomic Data Commons Data Portal](https://portal.gdc.cancer.gov). The data are also available through the R/Bioconductor packages [`curatedTCGAData`](https://www.bioconductor.org/packages/curatedTCGAData) and [`TCGABiolinks`](https://www.bioconductor.org/packages/TCGAbiolinks).

For simplicity, we have included the data in this repository as an `.RData` file.

We also include a tab-delimited file with some gene info useful for the GC-content normalization performed here.

```{r setup, message=FALSE}
jobName <- "LGG20190927"
library(awst)
library(EDASeq)
library(scales)
library(survival)
library(survminer)
library(GGally)
library(heatmap3)
library(dendextend)
library(knitr)
###############################################
annotation.df <- read.table("LGG_annotation.tsv", sep = "\t", row.names = 1)
annotation.col <- annotation.df[, grep(".col", colnames(annotation.df))]
colnames(annotation.col) <- gsub(".col", "", colnames(annotation.col) )
###############################################
save_plots <- FALSE
png_width_large <- 1500
png_height_large <- 750
png_width_small <- 700
png_height_small <- 400
png_res <- 1/300

source("functions.R")
```

```{r _GCcontFullQuantile20180124.RData, eval=TRUE, echo=FALSE}
if(!file.exists("Legacy_TCGA_LGG_GCcontFullQuantile.RData")) {
  load("Legacy_TCGA_LGG_Level3.RData")
  tmp <- strsplit(rownames(ddata), "\\|")
  geneNames <- matrix("", ncol = 2, nrow = length(tmp))
  rownames(geneNames) <- rownames(ddata)
  j <- 1 
  while(j <= length(tmp)) {
    geneNames[j, 1] <- tmp[[j]][1]
    geneNames[j, 2] <- tmp[[j]][2]
    j <- j + 1
  }
  rownames(ddata) <- geneNames[, 2]
  ###
  geneInfo <- read.csv("geneInfoHG19_20180124.tsv", sep = "\t", row.names = 1)
  length(duplicatedSymbol <- sort(names(which(table(geneInfo[, "hgnc_symbol"]) > 1))))
  wwhich <- which(rownames(ddata) %in% rownames(geneInfo))
  ddata <- ddata[wwhich,]
  geneInfo <- geneInfo[rownames(ddata),]
  sum(is.na(geneInfo[, "gcContent"]))
  ####
  ans <- newSeqExpressionSet(ddata, featureData = geneInfo)
  ans <- withinLaneNormalization(ans, "gcContent", which= "full", offset = TRUE)
  ans <- betweenLaneNormalization(ans, which = "full", offset = TRUE)
  ########################
  normCounts <- floor(exp(log(counts(ans) + 0.01) + offst(ans)) - 0.01)
  length(wwhich <- which(normCounts < 0))
  normCounts[wwhich] <- 0
  rownames(normCounts) <- geneInfo[rownames(normCounts), "hgnc_symbol"]
  entrez <- rownames(geneInfo)
  geneInfo <- cbind(entrez, geneInfo)
  rownames(geneInfo) <- geneInfo[, "hgnc_symbol"]
  ###
  save(normCounts, geneInfo, sample_map, file = "Legacy_TCGA_LGG_GCcontFullQuantile.RData")
} else {
  load("Legacy_TCGA_LGG_GCcontFullQuantile.RData")
}
```

```{r working.RData, echo=FALSE, include=FALSE, eval=TRUE}
if(!file.exists(paste0(jobName, "_dist_hclust.RData"))) {
exprData <- awst(normCounts, lambda = (lambda <- 13))
save(exprData, jobName, lambda, file = paste0(jobName, "_awst13.RData"))
dim(exprData <- gene_filter(exprData))#[1]   516 10212
save(exprData, jobName, lambda, file = paste0(jobName, "_working.RData"))
nrow_exprData <- nrow(exprData)
ncol_exprData <- ncol(exprData)
ddist <- dist(exprData)
save(ddist, nrow_exprData, ncol_exprData, jobName, lambda, file = paste0(jobName, "_dist.RData"))
hhc <- hclust(ddist, method = "ward.D2")
aCalinski <- calinski(hhc, gMax = 20)
save(hhc, aCalinski, nrow_exprData, ncol_exprData, jobName, lambda, file = paste0(jobName, "_dist_hclust.RData"))
} else {
  load(paste0(jobName, "_dist_hclust.RData"))
}
```

# Figure 4: the clustering

```{r dendrogram, fig.width=30, fig.height=15, echo=FALSE}
annotation.df <- annotation.df[hhc$labels,]
clustering.col <- annotation.col[hhc$labels,]
hhc$height <- hhc$height/max(hhc$height)

mmain  <-  paste0(jobName, " lambda = ", round(lambda), " (", nrow_exprData, " samples/", ncol_exprData, " genes)")
if(save_plots) {
  mmain <- ""
  png(paste0(jobName, "_plot_hclust_main.png"), width= png_width_large, height= png_height_large, res = png_res)
  }
plot(hhc, hang = -1, labels = FALSE, xlab = "euclidean/ward", sub = "", main = mmain)
###
obj <- aCalinski/max(aCalinski) * 0.5
G <- length(obj)
ccol <- rep("black", G)
for (g in 2:(G - 1)) {
  check <- obj[g - 1] < obj[g] & obj[g + 1] < obj[g]
  if (check) 
    ccol[g] <- "red"
}
xx <- floor(seq(from = 1, to = length(hhc$labels)*0.75, length.out = G))
nums <- paste(1:G); nums[1] <- ""
obj <- obj + 0.28
text(xx, obj, nums, col = ccol)
lines(xx, obj, col = "gray30", lty = "longdash")

clust.colorCode <- NULL
no_of_clusters <- 5
hh <- mean(c(hhc$height[length(hhc$height)-no_of_clusters+2], hhc$height[length(hhc$height)-no_of_clusters+1]))
abline(h = hh, col = "red")
clust5 <- clust5.col <- as.factor(cutree(hhc, k = no_of_clusters))
levels(clust5) <- paste0("c5", levels(clust5))
levels(clust5.col) <- c("red", "cyan", "sienna1", "blue", "magenta")
tt <- table(clust5, clust5.col)
colorCode <- colnames(tt)[apply(tt, 1, which.max)]
names(colorCode) <- rownames(tt)

clust.colorCode <- colorCode
annotation.df$clust5 <- clust5
annotation.df$clust5.col <- clustering.col$clust5 <- clust5.col
levels(clust5) <- c("c51_IDHwt", "c52", "c53_codel", "c54_nonCodel-pp", "c55_nonCodel-gp")
annotation.df$clust5.explanatory <- clust5

tt <- table(annotation.df$clust5.explanatory, annotation.df$clust5.col)
colorCode <- colnames(tt)[apply(tt, 1, which.max)]
names(colorCode) <- rownames(tt)
colorCode <- colorCode[-which(names(colorCode) %in% names(clust.colorCode))]
clust.colorCode <- c(clust.colorCode, colorCode)

tt <- table(annotation.df$clust5.explanatory)
llegend <- paste(names(tt), " (", tt, ")", sep = "")
ffill <- clust.colorCode[names(tt)]
legend(380, 0.95, legend = llegend, fill = ffill, y.intersp = 1, box.col = "white", border = "white", title = "clust5", title.adj = 0)

no_of_clusters <- 15
hh <- mean(c(hhc$height[length(hhc$height)-no_of_clusters+2], hhc$height[length(hhc$height)-no_of_clusters+1]))
clust5bis <- clust5bis.col <- as.factor(cutree(hhc, k = no_of_clusters))
levels(clust5bis) <- paste0("c5", levels(clust5bis))
levels(clust5bis)<- c("c51", "c52", "c531", "c532", "c541", "c55", "c541", "c533", "c541", "c55", "c55", "c542", "c51", "c51", "c52")
tmp <- as.factor(as.character(clust5bis)); names(tmp) <- names(clust5bis); clust5bis <- tmp
annotation.df$clust5bis <- clust5bis.col <- clust5bis
levels(clust5bis.col) <- c("red2", "cyan", "darkseagreen", "sienna1", "steelblue", "blue", "gold", "magenta")
annotation.df$clust5bis.col <- clustering.col$clust5bis <- clust5bis.col
levels(clust5bis) <- c("c51_IDHwt", "c52", "c531_O2", "c532_O3", "c533_O1", "c541_G-Cimp-high", "c542_G-Cimp-low", "c55_nonCodel-gp")
annotation.df$clust5bis.explanatory <- clust5bis
tt <- table(clust5bis, clust5bis.col)
colorCode <- colnames(tt)[apply(tt, 1, which.max)]
names(colorCode) <- rownames(tt)
colorCode <- colorCode[-which(names(colorCode) %in% names(clust.colorCode))]
clust.colorCode <- c(clust.colorCode, colorCode)
clustering.col <- clustering.col[, rev(colnames(clustering.col))]
colored_bars(colors = clustering.col, dend = as.dendrogram(hhc), y_scale = -0.15, y_shift = -0.16)
tt <- table(clust5bis)
llegend <- paste(names(tt), " (", tt, ")", sep = "")
legend(450, 0.95, legend = llegend, fill = clust.colorCode[names(tt)], y.intersp = 1, box.col = "white", border = "white", title = "clust5bis", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$exprClust.NEJM, annotation.df$exprClust.NEJM.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(1, 0.95, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "exprClust/NEJM", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$exprClust.CELL, annotation.df$exprClust.CELL.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(60, 0.95, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "exprClust/CELL", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$methClust.CELL, annotation.df$methClust.CELL.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(110, 0.95, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "methClust/CELL", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$CODEL.subtype, annotation.df$CODEL.subtype.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(160, 0.95, legend = gsub("kamoun_", "", rownames(tt)), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "CODEL subtype", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$original.subtype, annotation.df$original.subtype.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(210, 0.95, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "original subtype", title.adj = 0)
```

# Supplementary Figure 3: clustering on the residual features

Hierarchical clustering of the 516 LGG samples with the Ward’s linkage method from an Euclidean matrix involving 8,926 features having heterogeneity less than 0.10.

```{r _residual_dendrogram.pdf, fig.width=30, fig.height=15, echo=FALSE, eval=TRUE}
if(!file.exists("LGG20190927_residual_dist_hclust.RData")) {
  load("LGG20190927_working.RData")
  usedGenes <- colnames(exprData)
  load("LGG20190927_awst13.RData")
  dim(exprData <- exprData[, setdiff(colnames(exprData), usedGenes)])#[1]  516 8926
  nrow_exprData <- nrow(exprData)
  ncol_exprData <- ncol(exprData)
  ddist <- dist(exprData)
  save(ddist, nrow_exprData, ncol_exprData, jobName, lambda, file = paste0(jobName, "_residual_dist.RData"))
  hhc <- hclust(ddist, method = "ward.D2")
  aCalinski <- calinski(hhc, gMax = 20)
  save(hhc, aCalinski, nrow_exprData, ncol_exprData, jobName, lambda, file = paste0(jobName, "_residual_dist_hclust.RData"))
} else {
  load("LGG20190927_residual_dist_hclust.RData")
}
hhc$height <- hhc$height/max(hhc$height)
mmain <- paste0(jobName, " (", nrow_exprData, " samples/", ncol_exprData, " genes)")
if(save_plots) {
  mmain <- ""
  png(paste0(jobName, "_plot_hclust_residual.png"), width= png_width_large, height= png_height_large, res = png_res)
}
plot(hhc, hang = -1, labels = FALSE, xlab = "euclidean/ward", sub = "", main = mmain)
colored_bars(colors = clustering.col, dend = as.dendrogram(hhc), y_scale = -0.15, y_shift = -0.16)

tmp <- apply(tt <- table(annotation.df$exprClust.NEJM, annotation.df$exprClust.NEJM.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(90, 0.99, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "exprClust/NEJM", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$exprClust.CELL, annotation.df$exprClust.CELL.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(135, 0.99, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "exprClust/CELL", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$methClust.CELL, annotation.df$methClust.CELL.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(182, 0.99, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "methClust/CELL", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$CODEL.subtype, annotation.df$CODEL.subtype.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(230, 0.99, legend = gsub("kamoun_", "", rownames(tt)), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "CODEL subtype", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$original.subtype, annotation.df$original.subtype.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(270, 0.99, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "original subtype", title.adj = 0)
#
tt <- table(annotation.df$clust5.explanatory)
legend(330, 0.99, legend=names(tt), fill = clust.colorCode[names(tt)], y.intersp = 1, box.col = "white", border = "white", title = "clust5", title.adj = 0)
#
tt <- table(annotation.df$clust5bis.explanatory)
legend(385, 0.99, legend=names(tt), fill = clust.colorCode[names(tt)], y.intersp = 1, box.col = "white", border = "white", title = "clust5bis", title.adj = 0)
```

# Supplementary Figure 4

Kaplan-Meyer estimates of the survival curves associated with clust5 partition,
based on the overall survival time. The log-rank test returns p-value $\leq 2.2 \cdot 10^{-16}$.

```{r, fig.width=20, fig.height=11, fig.align = "center"}
tmp.df <- annotation.df
notIsNA <- which(!is.na(Surv(tmp.df$surv_time, tmp.df$surv_status)))
tmp.df <- tmp.df[notIsNA,]
coxph_H1 <- coxph(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5.explanatory, data = tmp.df)
if(save_plots) {
  mmain <- ""
  png(paste0(jobName, "_surv_clust5.png"), width= png_width_small, height= png_height_small, res = png_res)
  }
suppressWarnings(ggadjustedcurves(coxph_H1, data = tmp.df, individual.curves = TRUE, variable = "clust5.explanatory",  method = "marginal", palette = clust.colorCode[levels(tmp.df$clust5.explanatory)]) + xlab("time (months)"))
tmp <- survdiff(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5.explanatory, data = tmp.df)
names(tmp$n) <- gsub("clust5.explanatory=", "", names(tmp$n))
tmp
```

# Supplementary Figure 5

Kaplan-Meyer estimates of the survival curves associated with the two non-codel subsets c54 and c55,
based on the overall survival time. The log-rank test returns p-value = $0.2$.

```{r, fig.width=20, fig.height=11, fig.align = "center"}
tmp.df <- annotation.df
tmp.df <- tmp.df[tmp.df$clust5 == "c54" | tmp.df$clust5 == "c55",]
notIsNA <- which(!is.na(Surv(tmp.df$surv_time, tmp.df$surv_status)))
tmp.df <- tmp.df[notIsNA,]
tmp.df$clust5.explanatory <- factor(tmp.df$clust5.explanatory)
coxph_H1 <- coxph(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5.explanatory, data = tmp.df)
if(save_plots) {
  mmain <- ""
  png(paste0(jobName, "_surv_clust5nonCodel.png"), width= png_width_small, height= png_height_small, res = png_res)
  }
suppressWarnings(ggadjustedcurves(coxph_H1, data = tmp.df, individual.curves = TRUE, variable = "clust5.explanatory",  method = "marginal", palette = clust.colorCode[levels(tmp.df$clust5.explanatory)]) + xlab("time (months)"))
tmp <- survdiff(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5.explanatory, data = tmp.df)
names(tmp$n) <- gsub("clust5.explanatory=", "", names(tmp$n))
tmp
```

# Supplementary Figure 6

Kaplan-Meyer estimates of the survival curves associated with the subsets
c533/O1, c531/O2, and c532/O3 (Kamoun and others, 2016) in c53 (IDHmut-CODELS), based on the
overall survival time. The log-rank test returns p-value = 0.02.

```{r}
tmp.df <- annotation.df
tmp.df <- tmp.df[tmp.df$clust5.explanatory == "c53_codel",]
notIsNA <- which(!is.na(Surv(tmp.df$surv_time, tmp.df$surv_status)))
tmp.df <- tmp.df[notIsNA,]
tmp.df$clust5bis.explanatory <- factor(tmp.df$clust5bis.explanatory)
coxph_H1 <- coxph(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5bis.explanatory, data = tmp.df)
if(save_plots) {
  mmain <- ""
  png(paste0(jobName, "_surv_clust5bisCodel.png"), width= png_width_small, height= png_height_small, res = png_res)
  }
suppressWarnings(ggadjustedcurves(coxph_H1, data = tmp.df, individual.curves = TRUE, variable = "clust5bis.explanatory",  method = "marginal", palette = clust.colorCode[levels(tmp.df$clust5bis.explanatory)])+ xlab("time (months)"))
tmp <- survdiff(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5bis.explanatory, data = tmp.df)
names(tmp$n) <- gsub("clust5bis.explanatory=", "", names(tmp$n))
tmp
```

# Supplementary Figure 7 

Kaplan-Meyer estimates of the survival curves associated with the two subsets c541 (G-Cimp-High) ant c542 (G-Cimp-Low), based on the overall survival time. The log-rank test returns p-value = $2\cdot 10^{-4}$.

```{r}
tmp.df <- annotation.df
tmp.df <- tmp.df[tmp.df$clust5 == "c54",]
notIsNA <- which(!is.na(Surv(tmp.df$surv_time, tmp.df$surv_status)))
tmp.df <- tmp.df[notIsNA,]
tmp.df$clust5bis.explanatory <- factor(tmp.df$clust5bis.explanatory)
coxph_H1 <- coxph(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5bis.explanatory, data = tmp.df)
if(save_plots) {
  mmain <- ""
  png(paste0(jobName, "_surv_clust5bisGCIMP.png"), width= png_width_small, height= png_height_small, res = png_res)
  }
suppressWarnings(ggadjustedcurves(coxph_H1, data = tmp.df, individual.curves = TRUE, variable = "clust5bis.explanatory",  method = "marginal", palette = clust.colorCode[levels(tmp.df$clust5bis.explanatory)])+ xlab("time (months)"))
tmp <- survdiff(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5bis.explanatory, data = tmp.df)
names(tmp$n) <- gsub("clust5bis.explanatory=", "", names(tmp$n))
tmp
```

# Supplementary Figure 8

Kaplan-Meyer estimates of the survival curves associated with clust5bis partition,
based on the overall survival time. The log-rank test returns p-value  $\leq 2.2\cdot 10^{-16}$.

```{r, fig.width=20, fig.height=11, fig.align = "center"}
tmp.df <- annotation.df
notIsNA <- which(!is.na(Surv(tmp.df$surv_time, tmp.df$surv_status)))
tmp.df <- tmp.df[notIsNA,]
coxph_H1 <- coxph(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5bis.explanatory, data = tmp.df)
if(save_plots) {
  mmain <- ""
  png(paste0(jobName, "_surv_clust5bis.png"), width= png_width_small, height= png_height_small, res = png_res)
  }
suppressWarnings(ggadjustedcurves(coxph_H1, data = tmp.df, individual.curves = TRUE, variable = "clust5bis.explanatory",  method = "marginal", palette = clust.colorCode[levels(tmp.df$clust5bis.explanatory)]) + xlab("time (months)"))
tmp <- survdiff(Surv(tmp.df$surv_time, tmp.df$surv_status) ~ clust5bis.explanatory, data = tmp.df)
names(tmp$n) <- gsub("clust5bis.explanatory=", "", names(tmp$n))
tmp
```

# Session info

```{r}
sessionInfo()
```
