---
title: "Code to reproduce the analyses of the AWST paper: Analysis of the LGG data<br/> Figure 19 and 20."
author: "[S.M.Pagnotta](https://www.researchgate.net/profile/Stefano_Pagnotta) - Dept. of Science and Technology, Universita' degli Studi del Sannio (Benevento, Italy)</br> [D.Risso](https://www.researchgate.net/profile/Davide_Risso) - Dept. of Statistical Sciences, Universita' degli Studi di Padova (Padova, Italy)"
date:  'compiled on `r format(Sys.time(), "%B %d, %Y")`'
output: 
  html_document: 
    fig_height: 7
    fig_width: 7
    toc: yes
    toc_float: yes
    df_print: paged
fontsize: 11pt
editor_options: 
  chunk_output_type: console
---

```{r knitr setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=7, fig.align = "center") 
```

Here we show the code to reproduce the analyses of: Risso and Pagnotta (2020). Per-sample standardization and asymmetric winsorization lead to accurate classification of RNA-seq expression profiles. In preparation.

This file belongs to the repository: https://github.com/drisso/awst_analysis.

The code is released with license GPL v3.0.


In this script we analyze LGG data normalized with transformation of Townes et al. (2019).



```{r setup, message=FALSE}
rm(list = ls()) 
setwd("~/Dropbox/AWST/gitHub/lgg20200913_townes/")
jobName <- "LGG20200913"
library(awst)
library(EDASeq)
library(scales)
library(survival)
library(survminer)
library(GGally)
library(heatmap3)
library(dendextend)
library(knitr)
###
annotation.df <- read.table("https://raw.githubusercontent.com/drisso/awst_analysis/master/LowerGradeGliomaExperiment/LGG_annotation.tsv", sep = "\t", row.names = 1) #20200316
annotation.col <- annotation.df[, grep(".col", colnames(annotation.df))]
colnames(annotation.col) <- gsub(".col", "", colnames(annotation.col) )
###
save_plots <- FALSE
png_width_large <- 1500
png_height_large <- 750
png_width_small <- 700
png_height_small <- 400
png_res <- 1/300

source("https://raw.githubusercontent.com/drisso/awst_analysis/master/functions.R") #20200316

load("Legacy_TCGA_LGG_Level3.RData")  

features_with_those_selected_by_AWST.df <- read.table("../lgg20200316_AWST/features_with_those_selected.tsv", header = TRUE, row.names = 4)

AWST_clustering.df <- read.table("../lgg20200316_AWST/AWST_clustering.tsv", header = TRUE, row.names = 2)

#scryData <- read.table("./townes_LGG_scryData_pearson.tsv", header = TRUE, row.names = 1)
#scryData <- t(as.matrix(scryData))
#rownames(scryData) <- gsub("\\.", "-", rownames(scryData))
```



```{r scryData, echo=TRUE, include=FALSE, eval=TRUE}
if(!file.exists(paste0(jobName, "_scry.RData"))) {
  require(scry)
  scryData <- nullResiduals(ddata, type = "pearson")
  scryData <- t(scryData)
  save(scryData, file = paste0(jobName, "_scry.RData"))
} else load(paste0(jobName, "_scry.RData"))
wData <- scryData
```

\newpage
# clustering with 2500 HVG
```{r HVG2500}
what <- "HVG2500"
tmp <- apply(wData, 2, var)
tmp <- HVG2500 <- names(tail(sort(tmp), 2500))
exprData <-  wData[, tmp]
no_of_partitions <- c(4, 10)
```
```{r _plot_hclust_main.png, fig.width=30, fig.height=15, echo=FALSE}
nrow_exprData <- nrow(exprData)
ncol_exprData <- ncol(exprData)
ddist <- dist(exprData)
hhc <- hclust(ddist, method = "ward.D2")
aCalinski <- calinski(hhc, gMax = 20)
annotation.col <- annotation.col[hhc$labels,]
hhc$height <- hhc$height/max(hhc$height)

mmain  <-  paste0(jobName, " (", nrow_exprData, " samples/", ncol_exprData, " genes)")
if(save_plots & (what %in% c("HVG10212", "AWST10212"))) {
  mmain <- ""
  png(paste0(jobName, "_plot_hclust_VST_", what, ".png"), width= png_width_large, height= png_height_large, res = png_res)
  }
plot(hhc, hang = -1, labels = FALSE, xlab = "euclidean/ward", sub = "", main = mmain)
add_Calinski_curve(height = 0.5, from = 1, to = 400)
colored_bars(colors = annotation.col, dend = as.dendrogram(hhc), y_scale = -0.1, y_shift = -0.1)

tmp <- apply(tt <- table(annotation.df$exprClust.NEJM, annotation.df$exprClust.NEJM.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(1, 0.95, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "exprClust/NEJM", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$exprClust.CELL, annotation.df$exprClust.CELL.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(60, 0.95, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "exprClust/CELL", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$methClust.CELL, annotation.df$methClust.CELL.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(110, 0.95, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "methClust/CELL", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$CODEL.subtype, annotation.df$CODEL.subtype.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(160, 0.95, legend = gsub("kamoun_", "", rownames(tt)), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "CODEL subtype", title.adj = 0)
#
tmp <- apply(tt <- table(annotation.df$original.subtype, annotation.df$original.subtype.col), 1, which.max)
ffill <- colnames(tt)[tmp]
legend(210, 0.95, legend = rownames(tt), fill = ffill, 
       border = "white", y.intersp = 1, box.col = "white", title = "original subtype", title.adj = 0)
```

\newpage
# clustering with 5000 HVG
```{r HVG5000}
what <- "HVG5000"
tmp <- apply(wData, 2, var)
tmp <- HVG5000 <- names(tail(sort(tmp), 5000))
exprData <-  wData[, tmp]
```
```{r, ref.label="_plot_hclust_main.png", fig.width=30, fig.height=15, echo=FALSE}
```

\newpage
# clustering with 10212 HVG (same number of AWST protocol)
```{r HVG10212}
what <- "HVG10212"
tmp <- apply(wData, 2, var)
tmp <- HVG10212 <- names(tail(sort(tmp), sum(features_with_those_selected_by_AWST.df$included)))
exprData <-  wData[, tmp]
```
```{r, ref.label="_plot_hclust_main.png", fig.width=30, fig.height=15, echo=FALSE}
```

\newpage
# clustering with 10212 features selected with the AWST protocol
```{r AWST}
what <- "AWST10212"
tmp <- which(features_with_those_selected_by_AWST.df$included)
tmp <- AWST10212 <- rownames(features_with_those_selected_by_AWST.df)[tmp]
exprData <-  wData[, tmp]
```
```{r, ref.label="_plot_hclust_main.png", fig.width=30, fig.height=15, echo=FALSE}
```

\newpage
# Session info

```{r}
sessionInfo()
```


```{r}
knit_exit()
```
