---
title: "Code to reproduce the analyses of the AWST paper: Analysis of the Synthetic data<br/>Supplementary Table 4."
author: "[S.M.Pagnotta](https://www.researchgate.net/profile/Stefano_Pagnotta) - Dept. of Science and Technology, Universita' degli Studi del Sannio (Benevento, Italy)</br> [D.Risso](https://www.researchgate.net/profile/Davide_Risso) - Dept. of Statistical Sciences, Universita' degli Studi di Padova (Padova, Italy)"
date:  '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document: 
    toc: yes
    toc_float: yes
    df_print: paged
fontsize: 11pt
editor_options: 
  chunk_output_type: console
---

Here we show the code to reproduce the analyses of: Risso and Pagnotta (2020). Per-sample standardization and asymmetric winsorization lead to accurate classification of RNA-seq expression profiles. In preparation.

This file belongs to the repository: https://github.com/drisso/awst_analysis.

The code is released with license GPL v3.0.

```{r setup, echo=FALSE, message=FALSE}
rm(list = ls())
setwd("~/Dropbox/AWST/gitHub/Synthetic20201003/")
library(knitr)
opts_chunk$set(echo = FALSE, fig.height = 4.5, fig.width = 11, fig.align = "center")
library(cluster)
library(dendextend)
library(clues)
library(awst)
jobName <- "Rob20201014"
###
# get data from gitHub
base_path <- "https://raw.githubusercontent.com/drisso/awst_analysis/master/"
ttable <- read.csv(url(paste0(base_path, "SyntheticExperiment/synthetic20200413.tsv")), sep = "\t")
annotation.df <- ttable[, 1:3]; rownames(annotation.df) <- annotation.df$samples
ddata <- ttable[, -(1:3)]
rownames(ddata) <- annotation.df$samples
colnames(ddata) <- gsub("X", "", colnames(ddata))
ddata <- t(ddata)
feature_annotation <- read.csv(url(paste0(base_path, "SyntheticExperiment/synthetic20200413_features_annotation.tsv")), sep = "\t")
ddata <- ddata[paste(feature_annotation$EntrezID), ]
rm(ttable)
tt <- table(annotation.df$sample, annotation.df$sample.col)
clust.col <- colnames(tt)[apply(tt, 1, which.max)]
names(clust.col) <- rownames(tt)
###
png_width <- 600
png_height <- 300
png_res <- 1/300
save_png <- FALSE
x_legend <- 81
y_scale_ <- -0.1
y_shift_ <- -0.155
source(url(paste0(base_path, "functions.R")))
results <- matrix(NA, ncol = 8, nrow = 150)
colnames(results) <- c("lambda", "sigma", "n_bins", "ECA", "ECP", "ARI", "G", "n features") 
k <- 0
```

# Study of Lambda, given sigma and number of bins in gene_filter
## Lambda = 13 (default value; all features)
```{r}
ssigma <- 0.15
nn_bins <- Inf
exprData <- awst(ddata, lambda = (llambda = 13))
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L13.png")
xxlab <- paste0("lambda = ", llambda)
```

```{r plot}
hhc <- hclust(ddist <- dist(exprData), method = "ward.D2")
aCalinski <- calinski(hhc, gMax = 7)
hhc$height <- hhc$height/max(hhc$height)
if(save_png) png(file = png_name, width = png_width, height = png_height, res = png_res)
plot(hhc, hang = -1, labels = FALSE, sub = "", main = mmain, xlab = xxlab)
wwhere <- 5
hh <- mean(c(hhc$height[length(hhc$height)-wwhere+2], hhc$height[length(hhc$height)-wwhere+1]))
segments(1, hh, nrow(exprData), hh, col = "red")
clustering <- as.factor(cutree(hhc, k = wwhere))
names(clustering) <- hhc$labels
tt <- table(clustering, annotation.df$sample.col)
levels(clustering) <- colnames(tt)[apply(tt, 1, which.max)]
tmp.col <- cbind(as.character(clustering), as.character(annotation.df$sample.col))
colnames(tmp.col) <- c("predicted", "ground truth")
colored_bars(colors = tmp.col, dend = as.dendrogram(hhc), y_scale = y_scale_, y_shift = y_shift_)
add_Calinski_curve(from = 25, to = 120)
results[k <- k+1,] <- c(llambda, ssigma, nn_bins, get_metrics(annotation.df$sample.col, clustering, verbose = TRUE)[1:4], ncol(exprData))
```

## Lambda = 10 (all features)
```{r Lambda = 10}
exprData <- awst(ddata, lambda = (llambda = 10))
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L10.png")
xxlab <- paste0("lambda = ", llambda)
```
```{r, ref.label="plot"}
```

## Lambda = 7.5 (all features)
```{r Lambda = 7.5}
exprData <- awst(ddata, lambda = (llambda = 7.5))
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L7dot5.png")
xxlab <- paste0("lambda = ", llambda)
```
```{r, ref.label="plot"}
```

## Lambda = 5 (all features)
```{r Lambda = 5}
exprData <- awst(ddata, lambda = (llambda = 5))
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L5.png")
xxlab <- paste0("lambda = ", llambda)
```
```{r, ref.label="plot"}
```

## Lambda = 2.5 (all features)
```{r Lambda = 2.5}
exprData <- awst(ddata, lambda = (llambda = 2.5))
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L2dot5.png")
xxlab <- paste0("lambda = ", llambda)
```
```{r, ref.label="plot"}
```

## Lambda = 0 (all features; simmetric smoothing)
```{r Lambda = 0}
exprData <- awst(ddata, lambda = (llambda = 0))
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L0.png")
xxlab <- paste0("lambda = ", llambda)
```
```{r, ref.label="plot"}
```


# Study of number of bins in gene_filter, given lambda and sigma
## number of bins = 40 in gene_filter
```{r}
llambda <- 13
exprData <- awst(ddata)
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L0.png")
xxlab <- paste0("lambda = ", llambda)
exprData <- gene_filter(exprData, nBins = (nn_bins <- 40))
```
```{r, ref.label="plot"}
```

## number of bins = 20 in gene_filter
```{r}
exprData <- awst(ddata, lambda = llambda)
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L0.png")
xxlab <- paste0("lambda = ", llambda)
exprData <- gene_filter(exprData, nBins = (nn_bins <- 20))
```
```{r, ref.label="plot"}
```

## number of bins = 10 in gene_filter
```{r}
exprData <- awst(ddata, lambda = llambda)
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L0.png")
xxlab <- paste0("lambda = ", llambda)
exprData <- gene_filter(exprData, nBins = (nn_bins <- 10))
```
```{r, ref.label="plot"}
```

## number of bins = 5 in gene_filter
```{r}
exprData <- awst(ddata, lambda = llambda)
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L0.png")
xxlab <- paste0("lambda = ", llambda)
exprData <- gene_filter(exprData, nBins = (nn_bins <- 5))
```
```{r, ref.label="plot"}
```


# Study of Sigma, given lambda and number of bins in gene_filter
## sigma_0 = 0.30 
```{r}
ssigma <- 0.30
nn_bins <- Inf
exprData <- awst(ddata, sigma0 = ssigma)
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L0.png")
xxlab <- paste0("lambda = ", llambda)
```
```{r, ref.label="plot"}
```

## sigma_0 = 0.075
```{r}
ssigma <- 0.075
exprData <- awst(ddata, sigma0 = ssigma)
mmain <- paste0(nrow(exprData), " samples/", ncol(exprData), " genes")
png_name <- paste0(jobName, "_L0.png")
xxlab <- paste0("lambda = ", llambda)
```
```{r, ref.label="plot"}
```


# Supplementary table 4
```{r}
results <- results[!is.na(results[,1]),]
kable(results)
#library(xtable)
#print(xtable(results), include.rownames = FALSE)
```


# Session info

```{r}
sessionInfo()
knit_exit()
```
