---
title: "Code to reproduce the analyses of the AWST paper: Supplemetary figure 5"
author: "[S.M.Pagnotta](https://www.researchgate.net/profile/Stefano_Pagnotta) - Dept. of Science and Technology, Universita' degli Studi del Sannio (Benevento, Italy)</br> [D.Risso](https://www.researchgate.net/profile/Davide_Risso) - Dept. of Statistical Sciences, Universita' degli Studi di Padova (Padova, Italy)"
date:  '`r format(Sys.time(), "%d %B, %Y")`'
output: 
  html_document: 
    toc: no
    toc_float: yes
    df_print: paged
fontsize: 11pt
editor_options: 
  chunk_output_type: console
---

Here we show the code to reproduce the analyses of: Risso and Pagnotta (2020). Per-sample standardization and asymmetric winsorization lead to accurate classification of RNA-seq expression profiles. In preparation.

This file belongs to the repository: https://github.com/drisso/awst_analysis.

The code is released with license GPL v3.0.

```{r setup, echo=FALSE, message=FALSE}
rm(list = ls())
setwd("~/Dropbox/AWST/gitHub/SD20200413/")
library(knitr)
opts_chunk$set(echo = FALSE, fig.height = 4.5, fig.width = 11, fig.align = "center")
library(cluster)
library(dendextend)
library(clues)
library(awst)
require(ConsensusClusterPlus)
jobName <- "ARIcurves202000511"
###
# get data from gitHub
base_path <- "https://raw.githubusercontent.com/drisso/awst_analysis/master/"
ttable <- read.csv(url(paste0(base_path, "SyntheticExperiment/synthetic20200413.tsv")), sep = "\t",
                   stringsAsFactors = TRUE)
annotation.df <- ttable[, 1:3]; rownames(annotation.df) <- annotation.df$samples
ddata <- ttable[, -(1:3)]
rownames(ddata) <- annotation.df$samples
colnames(ddata) <- gsub("X", "", colnames(ddata))
ddata <- t(ddata)
feature_annotation <- read.csv(url(paste0(base_path, "SyntheticExperiment/synthetic20200413_features_annotation.tsv")), sep = "\t")
ddata <- ddata[paste(feature_annotation$EntrezID), ]
rm(ttable)
tt <- table(annotation.df$sample, annotation.df$sample.col)
clust.col <- colnames(tt)[apply(tt, 1, which.max)]
names(clust.col) <- rownames(tt)
#load("synthetic20200413.RData")
###
thisPalette <- c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99") # from ConsensusClusterPlus
png_width <- 640
png_height <- 480
png_res <- 1/300
save_png <- FALSE
x_legend <- 81
y_scale_ <- -0.1
y_shift_ <- -0.13
source(url(paste0(base_path, "functions.R")))
source("~/Dropbox/consensus/pkg/consensusFun20200415.R")
##########
amount_of_features <- 1:10
results <- data.frame(row.names = paste0(amount_of_features, "K"), 
                      amount_of_features = 1000 * amount_of_features, 
                      FPKM = NA, FPKMccp = NA, FPKMccpPAM = NA)
```

```{r FPKM, eval=FALSE}
eData <- get_FPKM(ddata, feature_annotation$GeneLength)
eData <- log2(1 + eData)
get_exprData <- function(eMatrix, K) {
  tmp <- apply(eMatrix, 2, var)
  tmp <- names(tail(sort(tmp), K))
  eMatrix <-  scale(eMatrix[, tmp])
  return(eMatrix)
}
get_hc <- function(eData) {
  hhc <- hclust(dist(eData), method = "ward.D2")
  hhc$height <- hhc$height/max(hhc$height)
  hhc$main <- paste0(nrow(eData), " samples/", ncol(eData), " genes")
  return(hhc)
}
get_ARI_and_plot <- function(hhc, png_filename = NULL) {
  if(!is.null(png_filename)) png(file = png_filename, width = png_width, height = png_height, res = png_res)
  plot(hhc, hang = -1, labels = FALSE, xlab = "", sub = "", main = hhc$main)
  wwhere <- 5
  hh <- mean(c(hhc$height[length(hhc$height)-wwhere+2], hhc$height[length(hhc$height)-wwhere+1]))
  abline(h = hh, col = "red")
  clustering <- as.factor(cutree(hhc, k = wwhere))
  names(clustering) <- hhc$labels
  levels(clustering) <- thisPalette
  tmp.col <- cbind(as.character(clustering), as.character(annotation.df$sample.col))
  colnames(tmp.col) <- c("clust", "sample")
  colored_bars(colors = tmp.col, dend = as.dendrogram(hhc), y_scale = y_scale_, y_shift = y_shift_)
  tmp.col <- as.data.frame(tmp.col)
  message("adjusted-Rand index (clues): ", ans <- round(adjustedRand(as.numeric(tmp.col$clust), as.numeric(tmp.col$sample))[2], 4))
  if(!is.null(png_filename)) dev.off()
#  return(ifelse(ans < 0, 0, ans))
  return(ans)
}

for(k in results$amount_of_features) {
  exprData <- get_exprData(eData, k)
  hhc <- get_hc(exprData)
  results$FPKM[floor(k/1000)] <- get_ARI_and_plot(hhc)
}
```
```{r FPKMccp, eval=FALSE}
set.seed(20200509)
get_hc <- function(eData) {
  ans <- ConsensusClusterPlus(t(eData), maxK = 5, reps = 1000)
  hhc <- ans[[5]]$consensusTree
  hhc$height <- hhc$height/max(hhc$height)
  hhc$main <- paste0(nrow(eData), " samples/", ncol(eData), " genes")
  return(hhc)
}
for(k in results$amount_of_features) {
  exprData <- get_exprData(eData, k)
  hhc <- get_hc(exprData)
  results$FPKMccp[floor(k/1000)] <- get_ARI_and_plot(hhc)
}
```
```{r FPKMccp_PAM, eval = FALSE}
set.seed(20200509)
get_hc <- function(eData) {
  ans <- ConsensusClusterPlus(t(eData), maxK = 5, reps = 1000, clusterAlg = "pam", distance = "euclidean")
  hhc <- ans[[5]]$consensusTree
  hhc$height <- hhc$height/max(hhc$height)
  hhc$main <- paste0(nrow(eData), " samples/", ncol(eData), " genes")
  return(hhc)
}

for(k in results$amount_of_features) {
  exprData <- get_exprData(eData, k)
  hhc <- get_hc(exprData)
  results$FPKMccpPAM[floor(k/1000)]  <- get_ARI_and_plot(hhc)
}
save(results, file = paste0(jobName, "_results.RData"))
```

# Supplementary figure 5
Study of FPKM pre-processing on synthetic data. For each amount of features, 1) a clustering procedure has been applied, then 2) we got the adjusted Rand index. Three clustering procedures are shown: black) hierarchical clustering with Ward's linkage and Euclidean distance, red) ConsensusClusterPlus with average linkage (inner and outer) and Pearson's correlation as distance, and blue) ConsensusClusterPlus with PAM and Euclidean distance.
```{r ARIcurves202000511_FPKM.png}
load("ARIcurves202000511_results.RData")
if(save_png) png(file = paste0(jobName, "_FPKM.png"), width = png_width, height = png_height, res = png_res)
plot(results$FPKM, type = "l", main = "adjusted Rand index curves", xaxt = "n", bty = "n",
     xlab = "amount of features (in thousands)", 
     ylab = "adjusted Rand index")
text(1:length(results$FPKM), -0.025, rownames(results))
points(results$FPKM, pch = 20)
lines(results$FPKMccp, col = "red")
points(results$FPKMccp, pch = 20, col = "red")
lines(results$FPKMccpPAM, col = "blue")
points(results$FPKMccpPAM, pch = 20, col = "blue")
legend(5, 1, legend=c("HC with Ward's linkage and Euclidean distance", 
                      "CCP with average linkage and Pearson's correlation", 
                      "CCP with PAM and Euclidean distance"),
       col=c("black", "red", "blue"), lty=1, cex=0.8,  box.col = "white")
if(save_png) dev.off()
kable(results)
```

```{r}
sessionInfo()
knit_exit()
```
